experDesign: helping performing experiments on batches

### 2021-04-21 paper.Rmd
Remove setup from the output
Add word document, update authors name, remove code

---
title: "experDesign: helping performing experiments on batches"
output: 
  md_document:
    preserve_yaml: true
authors:
  - name: "Lluís Revilla Sancho"
    orcid: 0000-0001-9747-2570
    affiliation: '1, 2'
  - name: "Juan José Lozano Salvatella"
    orcid: 0000-0001-9747-2570
    affiliation: '1'
  - name: "Azucena Salas Martínez"
    orcid: 0000-0001-9747-2570
    affiliation: '2'
affiliations:
  - name: "Centro de Investigación Biomédica en Red, Enfermedades Hepáticas y Digestivas"
    index: 1
  - name: "Institut d'Investigacions Biomèdiques August Pi i Sunyer, IDIBAPS"
    index: 2
date: Sys.Date()
tags:
  - R
  - Batch effect
  - experiment design
bibliography: paper.bib
---

@@

There are several techniques to identify and take into account batch effects when analyzing an experiment already measured [@leek2010].
However, if the source of them are not taken into account before measuring they can introduce a batch effect that could be impossible to remove.
Therefore, it would be better if the batch effects were avoided before carrying out the experiment.
Tacking into account how the process from the design to the measurement can introduce variability will help avoiding batch effects.

To prevent the apparition of batch effect after the design of the experiment there are two options: randomization and replication.
This techniques can prevent the batch effect to be able to compare between or within the desired variable.
One can do both of them but in any case it must take into account the blocking design from the beginning or they might further exacerbate the problem.
For instance, if one designs an experiment with cases and controls and the controls are measured in one batch and the controls in a different batch then the difference between them can be entirely confounded by a malfunction of the measurement machine in one batch [@chen2011].

However, samples can be properly randomized to minimize the batch effect by looking how the variables distribute on each batch, this is known as randomized block experimental design or stratified random sampling experimental design.

Replications consist on increasing the number of measurements with similar attributes.
Usually at least three samples are included for each condition of interest to be able to estimate the variation of the attributes.
If there is one extraction and then a sample is measured multiple times it is called a technical replicate.
Technical replicates help estimate the variation of the measurement method or the process.

@@

There are some tools to prevent generating a batch effect on the R language from multiple fields and areas, but specially on the biological research [@rcoreteam2014].
They have some caveats that limit their application to some cases and to our knowledge there hasn't been a comparison of these different methods.
Here we describe existing packages briefly:

@@

-   Recently, *Omixer*, a new package on [Bioconductor](https://bioconductor.org/packages/Omixer/), has been made available [@sinke2021].
    It tests if the random assignments are homogeneous transforming all variables to numeric and using the Kendall's correlation if there are more than 5 samples, otherwise using the Pearson chi-squared test.

@@

The package *experDesign* provides the function `design` to distribute the samples on multiple batches so that each variable is homogeneous in each batch.
It is similar to the *anticluster* method for maximum variance.
If the experiment is carried out on a specific spatial distribution the `spatial` function distributes the samples homogeneously also by position similar to *Omixer*.

In addition to distribute the samples on batches *experDesign* provides with tools to add technical replicates.
When the design is already done then the replicates are called technical replicates [@blainey2014].
Technical replicates are samples that are measured multiple times and allow to reduce the uncertainty of the measurement.
If the technical replicates are distributed on several batches they allow to measure the batch effect and thus minimize it.
To select the technical replicates needed and from which samples the function `extreme_cases` is provided.
For easier usage the `replicates` designs an experiment with the number of replicates per batch desired.

*experDesign* also provides several small utilities to make it easier design the experiment in batches.
For instance, it helps to calculate the samples to distribute on the minimum number of batches required with `sizes_batches`.

# Comparison

For completeness a description of the usage of the different software

```{r example-data}
data(survey, package = "MASS")
survey$ID <- seq_len(nrow(survey))
VoI <- c("Sex", "Smoke", "Age")
n_batch <- 3
iterations <- 1000
```

## OSAT

OSAT provides some template for some setups:

```{r osat}
library("OSAT")
gs <- setup.sample(survey, optimal = VoI)
gc <- setup.container(IlluminaBeadChip96Plate, n = n_batch, batch = 'plates')
gSetup <- create.optimized.setup(sample = gs, container = gc, nSim = iterations)
```

## anticlust

```{r anticlust}
library("anticlust")
anticlust_index <- anticlustering(
  survey[, VoI],
  K = n_batch,
  objective = "variance",
  method = "exchange",
  repetitions = iterations
)
anticlust_index <- anticlustering(
  survey[, "Age"],
  K = n_batch,
  objective = "variance",
  method = "exchange",
  repetitions = iterations
)
```

## Omixer

```{r Omixer}
library("Omixer")
layout <- data.frame(well = 1:96, 
    row = factor(rep(1:8, each=12), labels = LETTERS[1:8]),
    column = rep(1:12, 8), 
    plate  = 1,
    chip = as.integer(ceiling(rep(1:12, 8)/2)))
layout$chipPos <- ifelse(layout$column %% 2 == 0, 
                         as.numeric(layout$row) + 8,
                         layout$row)
layout2 <- layout
layout2$plate <- 2
layout3 <- layout
layout3$plate <- 3
plates <- rbind(layout, layout2, layout3)
Omixer_index <- omixerRand(survey, sampleId = "ID", 
    block = "block", iterNum = 10, wells = 96,
    plateNum = n_batch, randVars = VoI, layout = plates, techVars = NULL)
```

## experDesign

```{r experDesign}
library("experDesign")
experDesign_index <- design(survey[, VoI], size_subset = 96,
                  iterations = iterations)
```

## Summary

Last we can compare some of the solutions

```{r compare}
OSAT_index <- get.experiment.setup(gSetup)$plates
table(batch_names(experDesign_index), anticlust_index)
# OSAT omits one sample
table(batch_names(experDesign_index)[-137], OSAT_index$plates)
table(anticlust_index[-137], OSAT_index$plates)
```

>>>>

---
title: "experDesign: helping performing experiments on batches"
output: 
  word_document: 
    keep_md: true
  md_document:
    preserve_yaml: true
authors:
  - name: "Lluís Revilla Sancho"
    orcid: 0000-0001-9747-2570
    affiliation: '1, 2'
  - name: "Juan-José Lozano"
    orcid: 0000-0001-9747-2570
    affiliation: '1'
  - name: "Azucena Salas"
    orcid: 0000-0001-9747-2570
    affiliation: '2'
affiliations:
  - name: "Centro de Investigación Biomédica en Red, Enfermedades Hepáticas y Digestivas"
    index: 1
  - name: "Institut d'Investigacions Biomèdiques August Pi i Sunyer, IDIBAPS"
    index: 2
date: Sys.Date()
tags:
  - R
  - batch effect
  - experiment design
bibliography: paper.bib
---

@@

There are several techniques to identify and take into account batch effects when analyzing an experiment already measured [@leek2010].
Experiments that run over long periods of time or are run across different laboratories are highly likely to be susceptible and even smaller single-laboratory studies if spanned several days or include personnel changes may be affected.
However, if the source of them are not taken into account before measuring they can introduce a batch effect that could be impossible to remove.
Therefore, it would be better if the batch effects were avoided before carrying out the experiment.
Tacking into account the process from the design to the measurement can help to avoid batch effects.

Experiments should be designed to distribute batches and other potential sources of experimental variation across groups.
To prevent the apparition of batch effect after the initial design of the experiment there are two options: randomization and replication.
This techniques can prevent the batch effect and allow to compare between or within the desired variable(s).
One can do both of them but in any case it must take into account the blocking design from the beginning or they might further exacerbate the problem.

Randomization if done correctly helps reducing the variation across groups while replication let us be more confident of the result of the experiment.
For instance, if one designs an experiment with cases and controls and the controls are measured in one batch and the controls in a different batch then the difference between them can be entirely confounded by a malfunction of the measurement machine in one batch [@chen2011].
However, by looking how the variables distribute on each batch samples can be properly randomized to minimize the batch effect, this is known as randomized block experimental design or stratified random sampling experimental design.

Replications consist on increasing the number of measurements with similar attributes.
Usually at least three samples are included for each condition of interest to be able to estimate the variation of the attributes.
If there is one extraction and then a sample is measured multiple times it is called a technical replicate.
Technical replicates help estimate the variation of the measurement method and thus of the possible batch effect.

@@

There are some tools to prevent generating a batch effect on the R language from multiple fields and areas, but specially on the biological research [@rcoreteam2014].
They have some caveats that limit their application to some cases and to our knowledge there hasn't been a comparison of these different methods.
Here we describe the current packages briefly:

@@

-   Recently, *Omixer*, a new package on [Bioconductor](https://bioconductor.org/packages/Omixer/), has been made available [@sinke2021].
    It tests if the random assignments are homogeneous transforming all variables to numeric and using the Kendall's correlation if there are more than 5 samples, otherwise using Pearson's chi-squared test.

@@

The package ***experDesign*** provides the function `design` to distribute the samples on multiple batches so that each variable is homogeneous in each batch.
It is similar to the *anticluster* method for maximum variance.
If the experiment is carried out on a specific spatial distribution the `spatial` function distributes the samples homogeneously also by position similar to *Omixer*.

In addition to distribute the samples on batches ***experDesign*** provides with tools to add technical replicates.
When the design is already done then the replicates are called technical replicates [@blainey2014].
Technical replicates are samples that are measured multiple times and allow to reduce the uncertainty of the measurement.
If the technical replicates are distributed on several batches they allow to measure the batch effect and thus minimize it.
To select the technical replicates needed and choose from which samples they are done the function `extreme_cases` is provided.
For easier usage the `replicates` designs an experiment with the number of replicates per batch desired.

***experDesign*** also provides several small utilities to make it easier design the experiment in batches.
For instance, it helps to calculate the samples to distribute on the minimum number of batches required with `sizes_batches`.
